# Introduction {#intro}

What is machine learning? Essentially, machine learning teaches computer models to look for patterns or make predictions. This might sound like magic or it might seem complicated, but you can think of machine learning models as finding underlying formulas that the data come from. To solve for such formula, many, many mathematical calculations are involved. As we human beings are prone to mistakes, as long as we can identify a framework, we can give the framework and data to a computer model. It is best at repeating meticulous calculations to find a best guess based on our believes of the system and the data we observed. 

## Supervised Learning  

## Unsupervised Learning  

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center', include = F}
# You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

# Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```



```{r nice-tab, tidy=FALSE, include = F}
# Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
# You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].
```


