# Data  

When you have data, don't feel so rushed to jump into data analysis yet. Some steps can help you know your data better and, more than often, avoid problems down the road. 

## Data Exploratoary Analysis  

After you read in data, do some checks. Below we use the embedded `mtcars` as an example for illustration.    

```{r, warning = FALSE}
library(tidyverse)
data("mtcars")
str(mtcars)
```

As seen above, the command `str` allows you to see the variables and their types. For this dataset, all variables are numerical. If some are categorical but they should be numerical, make sure you transform them into the right type of data. (Sometimes numbers can be saved as characters, and the analysis would not be correct if the datatype remains as character.) 

Next, try to make some plots--typically histograms for numerical variables and barplots for categorical variables.  

```{r}
par(mfrow = c(3,3), mar = c(2,2,2,2))
for (i in 1:ncol(mtcars)){
  hist(mtcars[,i], main = paste0("Histogram of ", colnames(mtcars)[i]))
}
```

You can also look at the paired plots to see if two variables are too perfectly correlated, which could cause problems in regression models.  

```{r}
pairs(mtcars)
```


Next, take a look at the summary statistics.  

```{r}
for (i in 1:ncol(mtcars)){
  print(paste0("***** Summaries of ", colnames(mtcars)[i]," *****"))
  print(summary(mtcars[,i]))
}
```

The summaries can show, for example, ranges and means of the variables. They can give you a better understanding of where the values are and whether there might be data entry errors.  



## Data Requirement  


## Strategies for Missing Data  

### Removing observations  

If you have a large dataset and the percentage of missing data is small, you can simply ignore those observations. Note that you should check whether the pattern of missingness is "missing at random," meaning that the reason why certain observations are missing is pure chance. If, for example, some observations are missing because the pollution level is too high or too low and the detection is not ideal, then ignoring these observations could result in biased conclusions.  

### Imputation  

Imputation is a commonly used method for missing data. If linearity is plausible, you can use linear imputation. If some response variables are missing, some Bayesian method can fill in the values using posterior draws. (In other words, in the process of running the analysis, the algorithm can predict what the response variable should be and fill it in.) Imputation is a big tool set that, if needed, you should consult a more comprehensive resource. [provide suggestions]  


